---
title: "City of Raleigh Budget Sentiment Analysis"
author: "Jason Jones"
date: 2018-04-29T21:09:00-00:00
categories: ["R"]
tags: ["R Markdown", "budget", "sentiment"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

```

## Package Import
Load necessary packages and set one global option.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(pdftools)
library(tidytext)
library(knitr)
library(kableExtra)
```

## Retrieve File
Download the file from the City of Raleigh website, read that file in as a character vector, and delete the downloaded file from the directory. 

```{r message=FALSE, warning=FALSE}
download.file("https://cityofraleigh0drupal.blob.core.usgovcloudapi.net/drupal-prod/COR11/FY2018AdoptedBudget20160612.pdf",
              "FY2018AdoptedBudget.pdf",
              mode = "wb")

txt = pdf_text("FY2018AdoptedBudget.pdf")

unlink("FY2018AdoptedBudget.pdf")

```

## Create Data Frame
Create a page number character vector, create a data frame by binding the page number character vector with the extracted text, and finally "unnest" all of the page text into individual words.

```{r message=FALSE, warning=FALSE}
page = as.character(1:length(txt))

df = data.frame(cbind(page, txt))

budget_words = df %>%
  mutate(txt = as.character(txt)) %>%
  unnest_tokens(word, txt)

```

## Cleaning 
Remove stop words and save as clean object, join sentiment lexicon with clean object, and group the object by page and sentiments before summarising.

```{r message=FALSE, warning=FALSE}
cleaned = budget_words %>%
  anti_join(stop_words)

sentiment = cleaned %>%
  inner_join(get_sentiments("nrc"))

sent_count = sentiment %>%
  group_by(page, sentiment) %>%
  summarise(sent_count = n()) %>%
  ungroup() %>%
  mutate(page = as.integer(page))

```

## Visualize

```{r out.width="100%", echo=FALSE}
sent_count %>%
  mutate(sentiment = toupper(sentiment)) %>%
  ggplot(aes(page, sent_count, fill = sentiment)) +
  geom_col() +
  labs(x = "Budget Document Page #") +
  scale_fill_viridis_d("Sentiment") +
  theme(panel.background = element_blank(),
        axis.title.y = element_blank(),
        text = element_text(family = "Roboto"),
        legend.position = "bottom")
```

## Negative Word Table

```{r echo=FALSE}
negative = sentiment %>%
  filter(sentiment == "negative") %>%
  group_by(word) %>%
  summarise(word_count = n()) %>%
  ungroup() %>%
  top_n(10, word_count) %>%
  mutate(word = toupper(word))

kable(negative, col.names = c("Word", "Word Count")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

## Trust Word Table

```{r echo=FALSE}
trust = sentiment %>%
  filter(sentiment == "trust") %>%
  group_by(word) %>%
  summarise(word_count = n()) %>%
  ungroup() %>%
  top_n(10, word_count) %>%
  mutate(word = toupper(word))

kable(trust, col.names = c("Word", "Word Count")) %>%
  kable_styling(bootstrap_options = c("stripted", "hover"))

```

