---
title: "Spatial Exploration - MUSA Masterclass"
author: "Jason Jones"
date: 2018-11-18T21:09:00-00:00
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

This post was inspired by the [MUSA Masterclass](https://github.com/PennMUSA/masterclass2018) provided very graciously by [Ken Steif](http://urbanspatialanalysis.com/) and [James Cheshire](http://spatial.ly/). If you aren't familiar with the program I would strongly encourage you to check it out - [UPenn MUSA Program](https://www.design.upenn.edu/musa/post/spring-18-recap).

I decided to use open data from the City of Raleigh for my exploration. All data can be found on [The City of Raleigh's open data portal](http://data-ral.opendata.arcgis.com/). The primary data I am using is contained in the [Raleigh Police Incidents (NIBRS)](http://data-ral.opendata.arcgis.com/datasets/raleigh-police-incidents-nibrs) dataset. Each row represents a report made by a police officer.

## Load Required Packages
Trying to keep it as simple as possible while leveraging some tidyverse tricks I've been trying to learn. Forcing myself to use them seems to be a good idea.

```{r}
library(sf)
library(tidyverse)
library(RANN2)
```

## Retrieve Data
Would strongly recommend downloading this data and reading from your computer. Here I am reading the data directly from the GeoJSON API endpoints.

```{r}
raleigh_police <- read_sf("https://opendata.arcgis.com/datasets/24c0b37fa9bb4e16ba8bcaa7e806c615_0.geojson")

wake_police_dpts <- read_sf("https://opendata.arcgis.com/datasets/23094dc3a7b84682898c0a2c27290066_0.geojson")

raleigh_limits <- read_sf("https://opendata.arcgis.com/datasets/4303065aa95441308cc7224cf6246782_0.geojson") %>%
  filter(LONG_NAME == "RALEIGH")
```

## Filter Missing Data
To prevent some problems down the road, I am filtering out all incidents that are missing spatial attributes.

```{r}
raleigh_police <- raleigh_police %>%
  filter(is.na(latitude) != TRUE)
```

## Test Plots
Just throwing out some quick test plots to make sure everything is working as anticipated.

```{r}
plot(raleigh_police$geometry, col = "grey")

plot(wake_police_dpts$geometry, col = "grey")
```

## Check for matching projection
Checking for matching projections for plotting and k nearest neighbor clustering.

```{r}
st_crs(raleigh_police)

st_crs(wake_police_dpts)
```

## Create Nested Tibble
Here I am nesting the data by crime description so I can apply the clustering function. This creates a column of tibbles for each crime description which will become important as we try to distinguish densities in later steps.

```{r}
raleigh_police_nested <- raleigh_police %>%
  group_by(crime_description) %>%
  nest()
```

## Create Clustering Function
Next, I wrote a function to apply to my column of tibbles. I am using the `nn2` function from the `RANN2` package to perform k nearest neighbor clustering. This function clusters points by law enforcement stations.

```{r}
f <- function(dat) {
  nn2(st_coordinates(wake_police_dpts), st_coordinates(dat), k =1) %>%
    data.frame() %>%
    as_tibble() %>%
    group_by(nn.idx) %>%
    summarise(cnt = n()) %>%
    right_join(wake_police_dpts, by = c("nn.idx" = "OBJECTID")) %>%
    select(-nn.idx) %>%
    filter(is.na(cnt) != TRUE)
}
```

## Map Clustering Function to Tibble
Next, I leverage the `purrr` package to map the function to the column of tibbles to create a new column of tibbles containing counts by law enforcement stations.

```{r}
raleigh_police_nested <- raleigh_police_nested %>%
  mutate(closest = purrr::map(.x = data, .f = ~f(.x)))
```

## Define Crime Description
Now we are ready to explore the results. The thought here is potentially a `shiny` application where you could specify a crime description and produce a series of informative plots. I start by defining a new object, `crime_des`, to filter for only Traffic/DWI Incidents.

```{r}
crime_des <- "Traffic/DWI (Driving While Impaired)"
```

## Extract Nested Tibble
Here, I am creating two new objects for plotting based on the crime description I just specified.

```{r}
dat <- raleigh_police_nested %>%
  filter(crime_description == crime_des) %>%
  .$closest %>%
  .[[1]] %>%
  st_as_sf()

# Also creating secondary object for density plots
dat2 <- raleigh_police %>%
  filter(crime_description == crime_des) %>%
  st_coordinates() %>%
  as_tibble()
```

## Density Plot
Now for the fun! First is a density map of Traffic/DWI incidents.

```{r}
dat2 %>%
  ggplot() +
  geom_sf(data = raleigh_limits, color = "grey", fill = NA) +
  stat_density_2d(aes(X, Y, fill = ..level..), geom = "polygon", alpha = 0.6) +
  viridis::scale_fill_viridis(option = "magma", direction = -1, name = "Density") +
  labs(title = sprintf("Density of %s Incidents in Raleigh, NC", crime_des),
       caption = "Author: Jason Jones \nSource: http://data-ral.opendata.arcgis.com/") +
  theme(text = element_text(family = "Roboto"),
        plot.title = element_text(size = 14),
        panel.background = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank())
```

## Hexbin Plot
Next is a Hexbin map of Traffic/DWI Incidents.

```{r}
dat2 %>%
  ggplot() +
  geom_sf(data = raleigh_limits, color = "grey", fill = NA) +
  geom_hex(aes(X,Y), alpha = 0.6) +
  viridis::scale_fill_viridis(option = "magma", direction = -1, name = "Density") +
  labs(title = sprintf("Density of %s Incidents in Raleigh, NC", crime_des),
       caption = "Author: Jason Jones \nSource: http://data-ral.opendata.arcgis.com/") +
  theme(text = element_text(family = "Roboto"),
        plot.title = element_text(size = 14),
        panel.background = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank())
```

## Clustering By Police Stations
Finally, is a map of law enforcement stations where the count of clustered incidents is mapped to the size of the point.

```{r}
dat %>%
  st_coordinates() %>%
  as_tibble() %>%
  mutate(cnt = dat$cnt, AGENCY = dat$AGENCY) %>%
  ggplot() +
  geom_sf(data = raleigh_limits, color = "grey", fill = NA) +
  geom_point(aes(X, Y, size = cnt, color = AGENCY), alpha = 0.8) +
  scale_color_brewer(palette = "Spectral", direction = -1, name = "Agency") +
  scale_size(name = "Count of Incidents") +
  labs(title = sprintf("Clustering of %s Incidents by LE Stations", crime_des),
       caption = "Author: Jason Jones \nSource: http://data-ral.opendata.arcgis.com/") +
  theme(text = element_text(family = "Roboto"),
        plot.title = element_text(size = 14),
        panel.background = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank())
```

Maybe one more time just for fun!

## Define Crime Description Again

```{r}
crime_des <- "Assault/Simple"
```

## Extract Nested Tibble Again

```{r}
dat <- raleigh_police_nested %>%
  filter(crime_description == crime_des) %>%
  .$closest %>%
  .[[1]] %>%
  st_as_sf()

# Also creating secondary object for density plots
dat2 <- raleigh_police %>%
  filter(crime_description == crime_des) %>%
  st_coordinates() %>%
  as_tibble()
```

## Density Plot Again

```{r}
dat2 %>%
  ggplot() +
  geom_sf(data = raleigh_limits, color = "grey", fill = NA) +
  stat_density_2d(aes(X, Y, fill = ..level..), geom = "polygon", alpha = 0.6) +
  viridis::scale_fill_viridis(option = "magma", direction = -1, name = "Density") +
  labs(title = sprintf("Density of %s Incidents in Raleigh, NC", crime_des),
       caption = "Author: Jason Jones \nSource: http://data-ral.opendata.arcgis.com/") +
  theme(text = element_text(family = "Roboto"),
        plot.title = element_text(size = 14),
        panel.background = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank())
```

## Hexbin Plot Again

```{r}
dat2 %>%
  ggplot() +
  geom_sf(data = raleigh_limits, color = "grey", fill = NA) +
  geom_hex(aes(X,Y), alpha = 0.6) +
  viridis::scale_fill_viridis(option = "magma", direction = -1, name = "Density") +
  labs(title = sprintf("Density of %s Incidents in Raleigh, NC", crime_des),
       caption = "Author: Jason Jones \nSource: http://data-ral.opendata.arcgis.com/") +
  theme(text = element_text(family = "Roboto"),
        plot.title = element_text(size = 14),
        panel.background = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank())
```

## Clustering By Police Stations Again

```{r}
dat %>%
  st_coordinates() %>%
  as_tibble() %>%
  mutate(cnt = dat$cnt, AGENCY = dat$AGENCY) %>%
  ggplot() +
  geom_sf(data = raleigh_limits, color = "grey", fill = NA) +
  geom_point(aes(X, Y, size = cnt, color = AGENCY), alpha = 0.8) +
  scale_color_brewer(palette = "Spectral", direction = -1, name = "Agency") +
  scale_size(name = "Count of Incidents") +
  labs(title = sprintf("Clustering of %s Incidents by LE Stations", crime_des),
       caption = "Author: Jason Jones \nSource: http://data-ral.opendata.arcgis.com/") +
  theme(text = element_text(family = "Roboto"),
        plot.title = element_text(size = 14),
        panel.background = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank())
```

I really appreciate the class taught by James Cheshire even though this is a little bit of a deviation from clustering to roads. I am thinking something like this could be suited for a `shiny` application. Not sure if I will ever get around to making that happen.








