---
title: "Weather Data"
author: "Jason Jones"
date: 2019-05-29T21:13:00-00:00
output:
  html_document:
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.align = 'center', warning = FALSE, message = FALSE)
```

# The Boring Bits
```{r}
# Load packages ----
library(tidyverse)
library(tigris)
library(sf)
library(riem)
library(kableExtra)
```

<br>

# Finding Home

Let's start by grabbing the Chapel Hill boundary and then doing a test plot.
```{r results='hide'}
# Retrieve Chapel Hill boundary ----
chapel_hill <- places(state = 37, year = 2017) %>%
  st_as_sf() %>%
  filter(NAME == "Chapel Hill")
```

<br>

```{r}
chapel_hill %>%
  ggplot() +
  geom_sf(fill = NA, color = "Red") +
  theme(panel.background = element_blank())
```

<br>

# Weather Stations

Now we need to grab all of the weather stations for NC and do another test plot to make sure everything is working. We should see a nice collection of dots that looks like North Carolina!
```{r}
# Retrieve weather stations ----
stations <- riem_stations(network = "NC_ASOS") %>%
  st_as_sf(coords = c("lon", "lat"), crs = st_crs(chapel_hill))

stations %>%
  ggplot() +
  geom_sf() +
  theme(panel.background = element_blank())
```

<br>

# Subset Chapel Hill

Getting fancy now! We will do some geoprocessing to select only the weather station(s) that fall within the Chapel Hill boundary we grabbed first. A test plot should confirm that this worked for us.
```{r}
# Select station in Chapel Hill ---
selection <- st_intersection(stations, chapel_hill)

# Confirm location ----
stations %>%
  filter(id == "IGX") %>%
  ggplot() +
  geom_sf(data = chapel_hill, fill = NA, color = "Red") +
  geom_sf() +
  theme(panel.background = element_blank())
```

<br>

# Retrieve Weather Data

Now that we know what weather station we want to focus on, we can retrive all of the data for that station.
```{r}
# Retrieve weather data for station IGX ----
weather_dat <- riem_measures(station = "IGX")
```

<br>

UH-OH! Data shows that readings stopped in April 2018. A quick Google search revealed that the airport was actually closed in May 2018! Looks like we will need to try something else!

<br>

# Next Closest Station

Maybe we can do some more geoprocessing and find the next closest station. Let's start with that and then check out the results.
```{r}
# Compute distances of stations from Chapel Hill center
stations <- chapel_hill %>%
  st_centroid() %>%
  st_distance(stations, by_element = TRUE) %>%
  as.integer %>%
  cbind(stations) %>%
  rename(distance = ".")

# Check out the results
stations %>%
  as_tibble() %>%
  select(-geometry) %>%
  top_n(-3, distance) %>%
  arrange(distance) %>%
  kable(col.names = c("Distance", "Station ID", "Name")) %>%
  kable_styling()
```

<br>

# Weather Data One More Time

Looks like RDU is the next closest weather station to the center of Chapel Hill. Let's go get the weather data for that station now.
```{r}
# Retrieve weather data again ----
weather_dat <- riem_measures(station = "RDU")
```

<br>

Awesome stuff! Now we can play around with the weather data a bit.

# Monthly Average Temperature

```{r}
weather_dat %>%
  filter(is.na(tmpf) != TRUE) %>%
  mutate(valid = lubridate::floor_date(valid, unit = "month")) %>%
  group_by(valid) %>%
  summarise(avg_temp = mean(tmpf)) %>%
  ungroup() %>%
  ggplot(aes(valid, avg_temp)) +
  geom_line(color = "#112E51") +
  geom_point(size = 3, color = "#FF7043") +
  labs(x = NULL, y = NULL) +
  theme(panel.background = element_blank())
```

<br>

# Daily Highs and Lows
```{r}
highs <- weather_dat %>%
  filter(is.na(tmpf) != TRUE) %>%
  mutate(valid = lubridate::floor_date(valid, unit = "day")) %>%
  group_by(valid) %>%
  top_n(1, tmpf) %>%
  summarise(high = mean(tmpf)) %>%
  ungroup()

lows <- weather_dat %>%
  filter(is.na(tmpf) != TRUE) %>%
  mutate(valid = lubridate::floor_date(valid, unit = "day")) %>%
  group_by(valid) %>%
  top_n(-1, tmpf) %>%
  summarise(low = mean(tmpf)) %>%
  ungroup()

highs %>%
  left_join(lows, by = "valid") %>%
  gather("key", "value", 2:3) %>%
  ggplot(aes(valid, value, color = key)) +
  geom_line() +
  labs(x = NULL, y = NULL) +
  scale_color_manual(name = "Categories", values = c("#112E51", "#FF7043")) +
  theme(panel.background = element_blank())
```

<br>

# High and Low Differences, Monthly Variance
```{r}
highs %>%
  left_join(lows, by = "valid") %>%
  mutate(diff = high - low) %>%
  mutate(valid = lubridate::floor_date(valid, unit = "month")) %>%
  group_by(valid) %>%
  summarise(variance = var(diff)) %>%
  ggplot(aes(valid, variance)) +
  geom_line(color = "#112E51") +
  geom_point(size = 3, color = "#FF7043") +
  labs(x = NULL, y = NULL) +
  theme(panel.background = element_blank())
```

<br>
<br>