---
title: "Summer Meals: Census Tract Optimization"
author: "Jason Jones"
date: 2019-05-28T21:13:00-00:00
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

source("api_key/api_key.R")
```

# Required Packages

```{r}
library(tidyverse)
library(sf)
library(tidycensus)
library(sp)
library(tigris)
```


# Macro Geographies

## Durham County Polygon

```{r results='hide'}
nc_counties <- counties(state = "37")

durham_county_sf <- nc_counties %>%
  st_as_sf() %>%
  filter(NAME == "Durham")
```


## Durham County Census Tracts

```{r results='hide'}
durham_tracts <- tracts(state = "37", county = "063")

durham_tracts_sf <- durham_tracts %>%
  st_as_sf()
```


# Creating Decision Factors {.tabset .tabset-fade .tabset-pills}

To isolate the census tracts of potential focus, we will first use five different data points from ACS 5-years estimates for Durham County.

1. Population Age 3 And Over, Enrolled in School And In Poverty
2. Household Received Food Stamps/SNAP in the Past 12 Months
3. Median Household Income
4. Median Gross Rent as a Percentage of Household Income
5. Workers with No Vehicle Available

For factors 1, 2, 3, and 5 we will isolate census tracts by only selecting those that have estimates greater than the overall county median. For factor 4 we will simply select only census tracts that are greater than 30%.


## Decision Factor 1

```{r results='hide'}
school_age_poverty <- get_acs(geography = "tract", variables = "B14006_003", year = 2017, key = api_key,
                                state = "37", county = "063", geometry = TRUE, summary_var = "B14006_001", survey = "acs5")

pct_poverty <- school_age_poverty %>%
  mutate(pct = round(estimate/summary_est, digits = 2))

poverty_comare <- median(pct_poverty$pct, na.rm = TRUE)

pct_poverty %>%
  filter(pct > poverty_comare) %>%
  ggplot() +
  geom_sf(data = durham_county_sf, fill = "black", color = "black") +
  geom_sf(fill = "red") +
  theme(panel.background = element_blank())

df_1 <- pct_poverty %>%
  filter(pct > poverty_comare)
```


## Decision Factor 2

```{r results='hide'}
snap_recepients <- get_acs(geography = "tract", variables = "B22007_002", year = 2017, key = api_key,
                           state = "37", county = "063", geometry = TRUE, summary_var = "B22007_001")

pct_snap <- snap_recepients %>%
  mutate(pct = round(estimate/summary_est, digits = 2))

snap_compare <- median(pct_snap$pct, na.rm = TRUE)

snap_recepients %>%
  filter(estimate > snap_compare) %>%
  ggplot() +
  geom_sf(data = durham_county_sf, fill = "black", color = "black") +
  geom_sf(fill = "red") +
  theme(panel.background = element_blank())

df_2 <- snap_recepients %>%
  filter(estimate > snap_compare)
```


## Decision Factor 3

Pulling median for entire county because you should not compute median on a group of medians. Also, choosing for ACS 1-year estimate for entire county due to accuracy of estimate.

```{r results='hide'}
med_household_income <- get_acs(geography = "tract", variables = "B19013_001", year = 2017, key = api_key,
                                state = "37", county = "063", geometry = TRUE)

income_compare <- get_acs(geography = "county", variables = "B19013_001", year = 2017, key = api_key,
                          state = "37", county = "063", survey = "acs1") %>%
  .$estimate

med_household_income %>%
  filter(estimate < income_compare) %>%
  ggplot() +
  geom_sf(data = durham_county_sf, fill = "black", color = "black") +
  geom_sf(fill = "red") +
  theme(panel.background = element_blank())

df_3 <- med_household_income %>%
  filter(estimate < income_compare)
```


## Decision Factor 4

```{r results='hide'}
gross_rent <- get_acs(geography = "tract", variables = "B25071_001", year = 2017, key = api_key,
                      state = "37", county = "063", geometry = TRUE)

gross_rent %>%
  filter(estimate > 30) %>%
  ggplot() +
  geom_sf(data = durham_county_sf, fill = "black", color = "black") +
  geom_sf(fill = "red") +
  theme(panel.background = element_blank())

df_4 <- gross_rent %>%
  filter(estimate > 30)
```


## Decision Factor 5

```{r results='hide'}
no_vehicle <- get_acs(geography = "tract", variables = "B08014_002", year = 2017, key = api_key,
                      state = "37", county = "063", geometry = TRUE, summary_var = "B08014_001")

pct_no_vehicle <- no_vehicle %>%
  mutate(pct = round(estimate/summary_est, digits = 2))

vehicle_compare <- median(pct_no_vehicle$pct, na.rm = TRUE)
  
pct_no_vehicle %>%
  filter(pct > vehicle_compare) %>%
  ggplot() +
  geom_sf(data = durham_county_sf, fill = "black", color = "black") +
  geom_sf(fill = "red") +
  theme(panel.background = element_blank())

df_5 <- pct_no_vehicle %>%
  filter(pct > vehicle_compare)
```


# Putting It All Together

## Filtered Object

This creates an object containing only census tracts that meet all five decision factor criteria

```{r}
tracts_filt <- durham_tracts_sf %>%
  filter(GEOID %in% df_1$GEOID) %>%
  filter(GEOID %in% df_2$GEOID) %>%
  filter(GEOID %in% df_3$GEOID) %>%
  filter(GEOID %in% df_4$GEOID) %>%
  filter(GEOID %in% df_5$GEOID)
```


## Final Map Visual

```{r}
tracts_filt %>%
  ggplot() +
  geom_sf(data = durham_tracts_sf, fill = "light grey", color = "black") +
  geom_sf(fill = "red") +
  theme(panel.background = element_blank())
```


## Supplemental Table

```{r}
tracts_filt %>%
  as_tibble() %>%
  select(1:6) %>%
  kableExtra::kable() %>%
  kableExtra::kable_styling()
```



